Efficient implementation of penalized regression\\for genetic risk prediction

### 2019-02-08 paper/PRS.tex
save

{\color{red}
In this paper, we present an efficient method to jointly estimate SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals. Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters. We also provide a similar implementation of penalized linear regression.
}

We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
}
We find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR. We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.

{\color{red}
Using 350,000 individuals of the UK Biobank, we predict height with 10\% more correlation than with the best prediction of C+T. Although height is a highly polygenic trait with many small effects, our implementation of penalized regression proves to be scalable and, to the best of our knowledge, provides the best genetic predictor of height to date.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.
}

In conclusion, our study demonstrates that penalized regression can achieve much more predictive polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in R package bigstatsr.
}

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
Yet, PRS would have to show a high discriminative power between cases and controls in order to be used in clinical settings. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}. Note that these AUC figures are also dependent on the prevalence of the disease.
}

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR) against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For C+T, we report results for a large grid of hyper-parameters. 
For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation. We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data. %This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return. 
}

@@

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000. It compares all methods described in section \ref{sec:methods}. For each combination of the remaining parameters, results are based on 100 simulations except when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods. 
Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs). Reducing the number of SNPs increases the polygenicity (i.e.\ the proportion of causal SNPs) of the simulated models. Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power is dependent on $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}. For this scenario, we use the additive model only, but continue to vary all other simulation parameters. 
Finally, scenario \textnumero3 uses the whole dataset as in scenario \textnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.

\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
& & & 30 in all & & & & PLR \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
& & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
{\color{red}
For the computation of this paper, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
}

@@

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$.

To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \emph{linear} combination of allele counts. We refer to this method as ``PLR'' in the results section.

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of C+T and PLR in scenario \textnumero1 for model ``ADD''. Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$). Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-logit}
\end{figure}

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360.
In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds). 
{\color{red}
Finally, on this Celiac dataset, our implementation of PLR runs 300 times as fast as SparSNP \cite[]{abraham2012sparsnp}.
}

@@

We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing individuals related to someone else in the data (excluding the second individual in each pair with a kinship greater than 0.08). Results are presented in table \ref{tab:results-UKBB}.

Our L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes. The best C+T model achieves a correlation of 55\% for women and 56\% for men, where the GWAS part takes 1 hour only. 
If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
Our L1-penalized logistic regression on breast cancer runs in only 13 minutes for 150K women, achieving an AUC of 0.598. The best C+T model achieves an AUC of 0.589, where the GWAS part takes 15 hours.

\begin{table}[h]
\caption{Results for the UK Biobank. The training/test sets for height (resp. breast cancer) are 350,000/24,131 (resp. 150,000/38,628). For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set). Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set. For height, $r$ (correlation with true height) is reported within women / men; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|l|c|c|c|}
  \hline
Trait & Method & $r$ or AUC & \# predictors & Execution time \\
  \hline
height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
breast cancer & PLR & 0.598 & 2653 & 13m \\ 
breast cancer & C+T-max & 0.589 & 21 & 15h \\
  \hline
\end{tabular}
\end{table}

@@

The computation time of our PLR implementation mainly depends on the sample size and the number of active variables. Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes one statistic that will be used to decide if the variable enters the model or not (for each value of $\lambda$). This first step should be very fast.
Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}). So, in the second step, the number of variables is increasing progressively, and at some point the computation stops due to the early stopping criterion (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).

So, for highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over more than 100,000 variables, which takes time. On the contrary, for traits like the celiac disease or breast cancer that seems to be less polygenic, a lot less variables are entering the model and the fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer!). Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.

Regarding memory requirements, this is basically the same as for computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
If there are too many variables accessed, the operating system (OS) has to make some space for new coming variables. 
There should never be any memory issue thanks to memory-mapping. Yet, at some point, if there are too many active variables, the OS would regularly swap memory between disk and RAM, which is very slow. 
If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
We have quickly tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
Another widely used approach is to apply the penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset). Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper. Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}).

@@

When there are many effects of small size and a small sample size, PLR performs worse than (the best result for) C+T. This may be because PLR is too conservative in including effects, whereas it has been shown in the literature that, in such underpowered scenarios, including (nearly) all SNPs was the best strategy \cite[]{purcell2009common}.
In scenario \textnumero3, by varying sample size, we show that when increasing sample size, PLR achieves higher predictive performance than C+T at some point and that the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, in scenario \textnumero2, we decrease the search space (p: number of SNPs) in order to increase the polygenicity of the models and to virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}

@@

>>>>

{\color{red}
In this paper, we present an efficient method for the joint estimation of SNP effects using individual-level data, allowing for practical application of penalized logistic regression (PLR) on modern datasets including hundreds of thousands of individuals. Moreover, our implementation of PLR directly includes automatic choices for hyper-parameters. We also provide an implementation of penalized linear regression for quantitative traits.
}

We compare the performance of PLR, C+T and a derivation of random forests using both real and simulated data. 
{\color{red}
Overall, we find that PLR achieves equal or higher predictive performance than C+T in most scenarios considered, while being scalable to biobank data.
}
In particular, we find that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; for instance, in simulations, AUC values increase from 83\% with the best prediction of C+T to 92.5\% with PLR. We confirm these results in a data analysis of a case-control study for celiac disease where PLR and the standard C+T method achieve AUC of 89\% and of 82.5\%.

{\color{red}
Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T [TODO CHIFFRES FROM TO], further demonstrating its scalibility and strong predictive power even for highly polygenic traits.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.

In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an efficient implementation available in R package bigstatsr.
}
}

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact} [TODO: find prevalence].
}

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For C+T, we report results for a large grid of hyper-parameters. 
For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation. We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data. For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K).  %This demonstrates the feasibility and relevance of this approach for PRS computation based on large individual-level datasets that require scalable methods but offer some high predictive performance in return. 
}

@@

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes -- 281,122 SNPs) and a training set of size 6000. It compares all methods described in section \ref{sec:methods}. For each combination of the remaining parameters, results are based on 100 simulations except when comparing PLR with T-Trees, which relies on 5 simulations only because of a much higher computational burden of T-Trees as compared to other methods. 
Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only (18,941 SNPs). Reducing the number of SNPs increases the polygenicity (i.e.\ the proportion of causal SNPs) of the simulated models. Reducing the number of SNPs ($p$) is also equivalent to increasing the sample size ($n$) as predictive power increases as a function of $n/p$ \cite[]{dudbridge2013power,vilhjalmsson2015modeling}. For this scenario, we use the additive model only, but continue to vary all other simulation parameters. 
Finally, scenario \textnumero3 uses the whole dataset as in scenario \textnumero1 while varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen on the genome.

\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. [TODO: number of SNPs in dataset]\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & of training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
& & & 30 in all & & & & PLR \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
& & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients \[\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,\] where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only the SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 p-value thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.
{\color{red}
Here, the GWAS part uses the training set while clumping uses the test set (all individuals not included in the training set).
}

@@

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$, which is embedded in our PLR implementation.

To fit penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vector of estimated effect sizes is sparse. 
We refer to this method as ``PLR'' in the results section.

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'') [TODO: chnages for every plot]. Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$). Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-logit}
\end{figure}

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360.
In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds). [TODO: reformuler SparSNP reponse]

@@

We tested our implementation on 656K genotyped SNPs of the UK Biobank, keeping only caucasian individuals and removing related individuals (excluding the second individual in each pair with a kinship coefficient greater than 0.08). Results are presented in table \ref{tab:results-UKBB}.

Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes [TODO: IN.. ajouter test set]. By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men, and the GWAS part takes 1 hour. 
If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 [TODO test set]. The best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.

\begin{table}[h]
\caption{[TODO deplace dfmax supmat] Results for the UK Biobank. The training/test sets for height (resp. breast cancer) are 350,000/24,131 (resp. 150,000/38,628). For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set). Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set. For height, $r$ (correlation with true height) is reported within women / men; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|l|c|c|c|}
  \hline
Trait & Method & $r$ (women / men) or AUC & \# predictors & Execution time \\
  \hline
height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
breast cancer & PLR & 0.598 & 2653 & 13m \\ 
breast cancer & C+T-max & 0.589 & 21 & 15h \\
  \hline
\end{tabular}
\end{table}

@@

@@

When there are many small effects and a small sample size, PLR performs worse than (the best result for) C+T. For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K). 
In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \textnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}

@@

The computation time of our PLR implementation mainly depends on the sample size and the number of active variables (variables that are included in the gradient descent). Indeed, the algorithm is composed of two steps: first, for each variable, the algorithm computes an univariate statistic that is used to decide if the variable is included in the model (for each value of $\lambda$). This first step is very fast.
Then, the algorithm iterates over a regularization path of decreasing values of $\lambda$, which progressively enables variables to enter the model (Figure  \ref{fig:CMSA}). In the second step, the number of variables increases, and computations stop when an early stopping criterion is reached (when prediction is getting worse on the corresponding validation set, see figure \ref{fig:CMSA}).

For highly polygenic traits such as height and when using huge datasets such as the UK Biobank, the algorithm might iterate over more than 100,000 variables, which is computationally demanding. On the contrary, for traits like celiac disease or breast cancer that are less polygenic, the number of variables included in the model is much smaller so that fitting is very fast (13 minutes only for 150K women of the UK Biobank for breast cancer). %Using more cases (there were only 3542 for Celiac and 6768 for breast cancer), one might achieve better predictive performance for these two diseases, and the computation time should remain within a few hours.
